---
title: "Solutions to make your model converge"
author: "Heloise Thero"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Solutions to make your model converge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette concerns the case in which your model did not manage to converge, i.e., the Gelman-Rubin diagnostic stayed above 1.1 for at least one variable. There are a few solutions that can be tried.

Of course before running the model, we neet to have preprocessed the data and written the model:

```{r}
library(EcoDiet)
example_stomach_data_path <- system.file("extdata", "example_stomach_data.csv",
                                    package = "EcoDiet")
example_isotope_data_path <- system.file("extdata", "example_isotope_data.csv",
                                    package = "EcoDiet")
data <- preprocess_data(stomach_data_path = example_stomach_data_path, 
                        isotope_data_path = example_isotope_data_path,
                        trophic_enrichment_factor = c(0.8, 3.4),
                        literature_prior = FALSE)
model_string <- write_model(literature_prior = FALSE)
```

## Increase the number of iterations

The first thing that should be tried is to increase the number of iterations to run the model. The approximated probability distributions are bound to converge to their true distributions if enough iterations are given. 

So you should try to increase the `n_iter` argument, althouh this will likely takes several hours to run:
```{r, eval = FALSE}
mcmc_output <- run_model(textConnection(model_string), data, n_iter = 1e+08)
```

## Specify initial values for the model

The next other thing that can be tried is to initiate the variables, i.e., to define starting points from which to run the MCMC chains.

Here we are using the data to define some fixed initialization values that we know are close to the values we want the model to converge to. This way the model should be faster to converge because it is given a good start:

```{r, eval = FALSE}
inits <- initialize_model(data, ecodiet_example$topo_run)
```
```{r, eval = FALSE}
mcmc_output <- run_model(textConnection(model_string), data, inits = inits, n_iter = 1e+06)
```

## Rescale your stomachal data

If the occurrences in your stomachal data are too small, you may need to rescale them, i.e., to increase them using the number of full stomachs as a reference. For that you can use the `rescale_stomach_data` function.

Here are our example stomachal data before rescaling them:

```{r}
data$o
```

And here are the stomachal data after they have been rescaled:

```{r}
data$o <- rescale_stomach_data(data$o, data$nb_o)
data$o
```

## Further understanding the issue

If you have tried everything and that nothing works, it can be valuable to better understand the problem.

We are using the Gelman-Rubin statistic (or potential scale reduction factor) to measure convergence for each variable. The full statistic is saved as `convergence_diagnostic`, and you can acess it using:
```{r, eval = FALSE}
load("convergence_diagnostic.Rdata")
View(gelman)
```
It will allow you to understand which variables has not converged properly (the closer it is to 1, the better it has converged).

If you have a doubt about the relevance of the Gelman-Rubin statistic, or if you would like a second opinion about your convergence problem, you can also try the other convergence diagnostics available in the `coda` package.

## Environment on which this vignette has been run

```{r}
devtools::session_info()
```
