---
title: "EcoDiet explored on a realistic example"
author: "Heloise Thero"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. EcoDiet explored on a realistic example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

We will use a more realistic example (although still artificial!) to run the different EcoDiet configurations and compare their results.

## The configuration without literature data

```{r}
realistic_stomach_data_path <- system.file("extdata", "realistic_stomach_data.csv",
                                           package = "EcoDiet")
realistic_stomach_data <- read.csv(realistic_stomach_data_path)
knitr::kable(realistic_stomach_data)
```

```{r}
realistic_biotracer_data_path <- system.file("extdata", "realistic_biotracer_data.csv",
                                           package = "EcoDiet")
realistic_biotracer_data <- read.csv(realistic_biotracer_data_path)
knitr::kable(realistic_biotracer_data[c(1:3, 31:33, 61:63), ])
```

```{r, fig.height = 5, fig.width = 8, fig.align = "center"}
library(EcoDiet)

plot_data(biotracer_data = realistic_biotracer_data,
          stomach_data = realistic_stomach_data)
```

```{r}
literature_configuration <- FALSE

data <- preprocess_data(biotracer_data = realistic_biotracer_data,
                        trophic_discrimination_factor = c(0.8, 3.4),
                        literature_configuration = literature_configuration,
                        stomach_data = realistic_stomach_data)
```

As you can see the priors are non-informative:

```{r, fig.height = 5, fig.width = 8, fig.align = "center"}
plot_prior(data, literature_configuration)
```

```{r}
model_string <- write_model(literature_configuration = literature_configuration)

mcmc_output <- run_model(textConnection(model_string), data, nb_adapt = 1e1, nb_iter = 1e2)
```

You should try to run the model until it converges (it can take a few **hours**):
```{r, eval = FALSE}
mcmc_output <- run_model(textConnection(model_string), data, nb_adapt = 1e3, nb_iter = 1e4)
```

```{r, fig.height = 5, fig.width = 8, fig.align = "center"}
plot_results(mcmc_output, data)
```

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
plot_results(mcmc_output, data, pred = "Pout")
```

You can also precise for which variable and preys you want to plot the probability distributions:

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
plot_results(mcmc_output, data, variable = "PI", pred = "Pout", prey = c("Bivalves", "Worms"))
```

## The configuration with literature data

```{r}
realistic_literature_diets_path <- system.file("extdata", "realistic_literature_diets.csv",
                                               package = "EcoDiet")
realistic_literature_diets <- read.csv(realistic_literature_diets_path)
realistic_literature_diets[, -1] <- round(realistic_literature_diets[, -1], 2)
knitr::kable(realistic_literature_diets)
```
